```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 5)
```

# Effective Visualizations

Now that you know how to create graphics and visualizations in R, you are armed with powerful tools for scientific computing and analysis. With this power also comes great responsibility. Effective visualizations is an incredibly important aspect of scientific research and communication. There have been several books (see references) written about these principles. In class today we will be going through several case-studies trying to develop some expertise into making effective visualizations. 

## Worksheet

**The worksheet questions for today are embedded into the class notes.**

You can download this Rmd file [here](https://github.com/STAT545-UBC/Classroom/raw/master/cm013.Rmd)

Note, there will be very little coding in-class today, but I've given you plenty of exercises in the form of a supplemental worksheet (linked at the bottom of this page) to practice with after class is over.

## Resources

1. [Fundamentals of Data Visualization](https://serialmentor.com/dataviz/introduction.html) by Claus Wilke.

1. [Visualization Analysis and Design](https://www-taylorfrancis-com.ezproxy.library.ubc.ca/books/9780429088902) by Tamara Munzner.

1. [STAT545.com - Effective Graphics](https://stat545.com/effective-graphs.html) by Jenny Bryan.

1. [ggplot2 book](https://ggplot2-book.org) by Hadley Wickam.

1. [Callingbull.org](https://callingbull.org/tools.html) by Carl T. Bergstrom and Jevin West.

## Part 1: Warm-up and pre-test [20 mins]

### Warmup:

Write some notes here about what "effective visualizations" means to you. Think of elements of good graphics and plots that you have seen - what makes them good or bad? Write 3-5 points.

1. Colourful and luring; should look interesting
1. Sells itself; many words are boring
1. Relevant; visualizations should be relevant to the main prose description
1. Do not make it too crowded. Spaces sometimes speak for themselves
1. Consider there can be colour-blind audience. Careful when choosing the colors

### CQ01: Weekly hours for full-time employees

> Question: Evaluate the strength of the claim based on the data: "German workers are more motivated and work more hours than workers in other EU nations."
>
> Very strong, strong, weak, very **week**, do not know

- Weak. Of course it *seems* that German workers work more hours on average. However, we need deviation to say anything about a comparison to other EU nations, let alone the argument about motivation. Besides, as the axis starts at 36, not the zero point, the differences seem to have be exaggerated.

- Main takeaway: Provide information about deviation. Start the axis at 0. 


### CQ02: Average Global Temperature by year

> Question: For the years this temperature data is displayed, is there an appreciable increase in temperature?
> 
> Yes, No, Do not know

- Do not know. First of all, appreciable increase is a vague term. Moreover, the trendline is difficult to follow

- Main takeaway: in this case, the y-axis should better be calibrated so that a change of a single degree can be noticeable.

### CQ03: Gun deaths in Florida

> Question: Evaluate the strength of the claim based on the data: “Soon after this legislation was passed, gun deaths sharply declined."
>
> Very strong, strong, weak, very **week**, do not know

- Very weak. Why is the y-axis reversed? 

- Main takeaway: Reversing the axis can mislead the reader. Besides, careful with colours.

## Part 2: Extracting insight from visualizations  [20 mins]

Great resource for selecting the right plot: https://www.data-to-viz.com/ ; encourage you all to consult it when choosing to visualize data.

### Case Study 1: Context matters
The scale for the y-axis is 0 - 0.6% on the left-hand side, while on the right-hand side, it is 86 - 95%.

### Case Study 2: A case for pie charts

## Part 3: Principles of effective visualizations [20 mins]

We will be filling these principles in together as a class

1.
1.
1.
1.
1.

### Make a great plot worse

Instructions: Here is a code chunk that shows an effective visualization. First, copy this code chunk into a new cell. Then, modify it to purposely make this chart "bad" by breaking the principles of effective visualization above. Your final chart still needs to run/compile and it should still produce a plot. 

How many of the principles did you manage to break?

## Plotly demo [10 mins]

Did you know that you can make interactive graphs and plots in R using the plotly library? We will show you a demo of what plotly is and why it's useful, and then you can try converting a static ggplot graph into an interactive plotly graph.

This is a preview of what we'll be doing in STAT 547 - making dynamic and interactive dashboards using R!


```{r}
library(tidyverse)
library(gapminder)
library(plotly)

p <- ggplot(gapminder, aes(x=gdpPercap, y=lifeExp, color=continent)) + 
  geom_point()

p %>% 
  ggplotly() # interactive

#plot_ly syntax
gapminder %>% plot_ly(x= ~gdpPercap, y= ~lifeExp, 
                      color= ~continent,
                      type='scatter',
                      mode='markers') 
```


## Supplemental worksheet (Optional)

You are highly encouraged to the cm013 supplemental exercises worksheet. It is a great guide that will take you through Scales, Colours, and Themes in ggplot. There is also a short guided activity showing you how to make a ggplot interactive using plotly.

- [Supplemental Rmd file here](https://github.com/STAT545-UBC/Classroom/blob/master/tutorials/cm013-supplemental.Rmd)
